Who does big team science?

### 2023-08-29 manuscript_scopus.Rmd
updating manuscript

  Thank you to Dwayne Lieck for providing an extensive list of large scale projects for this manuscript. 
abstract: |
  This paper examined the nature of publications in Big Team Science (BTS) - large-scale collaborations between multiple researchers at multiple institutions. As interest in BTS increases, it is useful to explore who is currently involved in BTS projects to determine diversity in both research subject and researcher representation. The types of publication outlets, number of publications, and subject areas of publication are presented to summarize the publications in BTS. Information about authors included in BTS will be presented including career length, numbers of publications/impact variables, education, and affiliation. Last, we will explore the representation of geopolitical regions by examining affiliation location to explore the impact of BTS on the de-WEIRD movement to diversify researcher representation. REWRITE THIS

@@

According to the Oxford English dictionary, collaboration is two or more
people working together to achieve a certain goal [@oed2016].
Collaboration in scientific endeavors involves multiple researchers at
(potentially) multiple institutions to communicate and work together to
advance knowledge in their chosen field. Collaboration can manifest
uniquely in each project dependent on the skill sets, hypotheses, and
perspectives of collaborators. While collaboration is not new in
science, the current interest of "big team science" is increasing
[@forscher2020; @stewart2017; @coles2022]. Big team science projects
and/or organizations utilize and run on large-scale collaboration to
ensure that diverse populations and ideas are brought into research
projects, which in turn allows for more reliability and generalizability
in the results and method of the study. For this study, Big Team Science
(BTS) will be defined as a collaboration of ten or more authors from at
least ten different institutions.

BTS appears to be increasing as a result of two sources: 1) increasing
globalization and technology that allows for real-time interdisciplinary
research, and 2) increasing interest in reproducibility, replication,
and generalizability [@maxwell2015; @nelson2018; @zwaan2018].
Technological advances have provided easier ways to collaborate with
people who are from other universities and countries through document
sharing platforms (e.g., Google, GitHub, and the Open Science
Framework), video chatting platforms (e.g., Zoom, Microsoft Teams), and
messaging and project management platforms (e.g., Slack, Trello,
when2meet, etc.). The credibility movement seems to suggest that by
having both collaborations that span across the globe and subfields of
research areas, age groups, and education levels should help to drive
science in the path of better materials, reliability, generalizability
and more robust sample sizes (when necessary) in a study
[@auspurg2021has; @nosek2014method; @lebel2018].

The credibility movement was originally defined by a focus on large
scale replications using in collaborative environments [@vazire2022].
Generally, the movement appears to have been driven by early career
researchers (i.e., those who are within five years of their first
appointment) [@maizey2019]; however, there are no large meta-scientific
investigations on this specific topic to date. Potentially, the lack of
investigation is tied to the newness of the large-scale research in many
fields, as it is only in recent years that publications like the Open
Science Collaboration [@opensciencecollaboration2015], Many Labs
Collaborations [for example, @klein2018; @ebersole2020; @ebersole2016;
@klein2022; @mathur2020; @buttrick2020; @skorb2020] or the first papers
from the Psychological Science Accelerator [@moshontz2018; @jones2021;
@dorison2022; @wang2021; @legate2022; @bago2022]. Generally, the
researcher incentive for replication was low: journals often prioritize
"novel" or new results which led to rejection of replication manuscripts
and publication bias [@hubbard1997; @franco2014; @nosek2012], the
"failure" to replicate was often placed on the replication team as "bad
science" rather than a careful consideration of publication biases and
(potential) questionable research practices [@ioannidis2015;
@maxwell2015; @klein2022], and why should someone want to spend time and
resources on an answer we already "know" [@isager2021; @isager2021a]?

@@

The credibility movement has been mirrored by the calls for
diversification or de-WEIRDing (e.g., Western, Educated, Industrialized,
Rich, and Democratic) scientific research [@henrich2010; @newson2021;
@rad2018] by improving representation in research samples. Like the
large-scale studies in Physics [@aphilos2021; @castelnovo2018] and
Biology [@collins2003], the social sciences struggle to represent the
breadth of humanity across both researcher and population
characteristics. Now, grassroots organizations, such as the
Psychological Science Accelerator [@moshontz2018], ManyBabies
(<https://manybabies.github.io/>), NutNet (<https://nutnet.org/>), and
DRAGNet (<https://dragnetglobal.weebly.com/>) can begin to tackle these
issues by recruiting research labs from all over the globe to provide
diversity in geographic, linguistic, and researcher representation.
Publications have examined the global understanding of morality, face
processing, COVID-19 information signaling, and more [@dorison2022;
@wang2021; @legate2022; @bago2022; @jones2021; @vanbavel2022]. While
these organizations and one-time groups for BTS studies have provided an
incredible wealth of data for the scientific community, we do not yet
know exactly *who* is involved with, and benefits from, the BTS and
credibility movement. Publications on BTS generally explore challenges,
lessons learned, and the need for BTS [@forscher2020; @coles2022].

@@

```{r fig-pub-time, include = TRUE, fig.cap="Number of big-team science publications separated by four large subject areas across years.", fig.width=10, fig.height=6}
ggplot(DF$pub_time_count %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(Year < 2023), 
       aes(Year, count, color = SubjectArea)) + 
  theme_classic() + 
  geom_point() + 
  scale_color_brewer(palette = "Spectral", name = "Subject Area") + 
  ylab("Number of Articles Per Year") + 
  xlab("Publication Year") + 
  theme(legend.position = "bottom")

@@

Publication interest area was summarized by the four large subject areas
creating a word cloud plot of the total number of publications within
the ASJCs. Figure \@ref(fig:fig-clouds) displays that the health
sciences tends to publish within medicine and oncology, with a
corresponding focus of cancer research and genetics for the life
sciences. The physical sciences is mostly dominated by physics research,
chemistry, and ecology. The BTS publications in the social sciences are
mostly within psychology, education, and health.

@@

The total number of unique authors across all publications was
`r total_authors`. The mean number of authors per publication was *M* =
`r avg_authors` (*SD* = `r sd_authors`, *Med* = `r med_authors`) with a
range of `r min_authors` to `r max_authors`. The median and average
number of authors by subject area are displayed in Figure
\@ref(fig:fig-author-year). In general, the average and median number of
authors increased over time, with the exception of the skew in the
physical sciences. Interestingly, the effect in the physical sciences
appears to be declining toward the general trends seen in other areas in
the last few decades.

@@

All values for this analyses were different from zero. The slopes for
the average career length were negative for all four subject areas,
indicating a trend toward younger scientist involvement over time for
each area, with the strongest effect in the Physical sciences. The
coefficient for variability in career length was also negative for each
of the four subject areas with the highest in the Physical sciences and
lowest in the Life Sciences. This result indicates a decrease in the
variability of career lengths over time, likely from two sources: 1)
more publications with more authors, thus, lowering variance
estimations, and 2) more young scholars overall. The effect sizes for
this analysis were surprisingly large ranging from $R^2$ to .25 to .47.
All values and their confidence intervals can be found on our OSF page.

@@

We used the same analyses described in the career length section to
analyze trends over time. An increasing slope over time indicates that
individuals who are publishing more are more represented in BTS over
time (i.e., increasing numbers of scholars with higher publication
rates), while a negative slope indicates more researchers with less
publications. A positive slope for the standard deviation of publication
metrics indicates increasing variance over time (i.e., more diversity in
the individual publication rates), while a negative slope would indicate
less diversity in researchers over time. While publication rates do not
represent value as a researcher, they are often used in hiring and
promotion decisions, and we used this variable as a proxy to gauge the
diversity in scholars represented in big teams. As shown in Figure
\@ref(fig:fig-heatmap) publication metrics were generally negative for
the average publication metrics, indicating more scholars over time with
lower numbers of publications with the strongest effects in health and
social sciences. The variability of publication counts was not
significant for the life sciences but was negative for the physical
sciences (less variability over time) and positive for social and health
sciences (more variability and over time). This result indicates that
the physical sciences are trending toward scholars with less
publications but also less diverse in number of publications, while the
health and social sciences see more diversity in publication counts and
less published scholars overall.

@@

Author geopolitical region is displayed in Figure
\@ref(fig:fig-map-both). Big team publications appear to be lead by
North America and Western Europe, while all publications are lead by
North America and East Asia. To understand the change in representation
diversity, we examined if the number of regions in a publication is
predicted by the year of publication. Increasing diversity would be
represented by a positive slope, while decreasing diversity would be
represented by a negative slope. As shown in Figure
\@ref(fig:fig-heatmap), the physical sciences do not show a trend of
change in representation, while all other sciences showed a positive
effect increasing in the number of geopolitical regions authors
represent on publications.

@@

In this investigation, we explored the publication rates, areas, and
researchers involved in big team science publications. Over a
half-million articles were published in nearly 15,000 journals since
1970 that qualified as big team science articles with at least ten
authors and ten different affiliations. The areas of publication aligned
to cancer and genetics research in medicine and oncology for medical and
life sciences, physics and chemistry for the physical sciences, and
psychology for the social sciences. All areas of research show an
explosion in growth in the number of publications and the number of
author included on manuscripts, replicating previous investigations into
this topic area [@sinatra2015; @wuchty2007; @hunter2008].

@@

The variability in the types of researchers involved in publications
also decreased across time in most areas of science with a decrease in
variability for career length. As mention, an increase in early career
researchers and numbers of publications could explain this effect
mathematically, potentially with other social influences mentioned
above. The variability in number of publications is decreasing in the
physical sciences, mirroring the career length results, but the opposite
effect was found in the health and social sciences. We see no clear
reason why career variability would decrease while the variability in
the number of publications would increase. The effect sizes for career
length were much larger than the effects for number of publications. One
speculation is the increasing requirements for a competitive faculty
role application. Given the limited number of positions, one potential
way to distinguish their application would be a larger number
publications in their early career [@caplow2017; @kyvik2003].

The number of geopolitical entities for researcher affiliation is
increasing over time, showing the results of globalization and the
ability to connect across time zones and cultures [@xie2014]. While our
definition of big team science required at least ten different
institutional affiliations, we did not filter papers by geopolitical
region, and thus, a manuscript could rely solely on institutions within
a single country. The physical sciences did not show an increase in
diversity of regions represented, however, it could be argued that the
development of large research centers like CERN forced earlier diversity
than other sciences (i.e., because CERN specifically recruited
scientists from sponsoring nations). The life, health, and social
sciences saw an increase in the number of regions represented with the
highest increase in the social science. This result likely corresponds
with an increased interest in big team science publications in
psychology [@coles2022; @forscher2020], and the desire to diversify the
populations represented in psychological research [@henrich2010;
@newson2021].

>>>>

  Thank you to Dwayne Lieck for providing an extensive list of large scale projects for this manuscript. 
abstract: |
  This paper examined the nature of publications in Big Team Science (BTS): large-scale collaborations between multiple researchers at multiple institutions. These projects can improve research by initiating collaborations that span across the globe, age groups, education levels, and subfields of research. As the number of BTS publications increase, it is useful to explore who is currently involved in BTS projects to determine diversity in both research subject and researcher representation. We examined the diversity of BTS publications and authors across more than half a million articles to investigate where and what is currently published, and author characteristics including differences in career length, publication metrics, affiliation, and affiliation geopolitical regions. Interestingly, BTS publications are increasingly dominated by early career researchers from WEIRD geopolitical regions with Health and Physical Science accounting for the majority of BTS articles. However, the increase in preprints, BTS articles, and non-WEIRD authors across time demonstrate the efforts of the science community to diversify its researchers.

@@

According to the Oxford English dictionary, collaboration is two or more
people working together to achieve a certain goal [@oed2016].
Collaboration in scientific endeavors involves multiple researchers at
(potentially) multiple institutions to communicate and work together to
advance knowledge in their chosen field. Collaboration can manifest
uniquely in each project dependent on the skill sets, hypotheses, and
perspectives of collaborators. While collaboration is not new in
science, the current interest of "big team science" is increasing
[@forscher2020; @stewart2017; @coles2022]. Big team science projects
and/or organizations utilize and run on large-scale collaboration to
ensure that diverse populations and ideas are brought into research
projects, which in turn allows for more reliability and generalizability
in the results and method of the study.

BTS appears to be expanding as a result of two sources: 1) increasing
globalization and technology that allows for real-time interdisciplinary
research, and 2) expanding interest in reproducibility, replication,
and generalizability [@maxwell2015; @nelson2018; @zwaan2018].
Technological advances have provided easier ways to collaborate with
people who are from other universities and countries through document
sharing platforms (e.g., Google, GitHub, and the Open Science
Framework), video chatting platforms (e.g., Zoom, Microsoft Teams), and
messaging and project management platforms (e.g., Slack, Trello,
when2meet, etc.). The credibility movement seems to suggest that by having both collaborations that span across the globe and subfields of research areas, age groups, and education levels should help to drive science in the path of better materials, reliability, generalizability and more robust sample sizes (when necessary) in a study [@auspurg2021has; @nosek2014method; @lebel2018].

The credibility movement was originally defined by a focus on large scale replications used in collaborative environments [@vazire2022]. Generally, the movement appears to have been driven by early career
researchers (i.e., those who are within five years of their first
appointment) [@maizey2019]; however, there are no large meta-scientific
investigations on this specific topic to date. Potentially, the lack of
investigation is tied to the newness of the large-scale research in many
fields, as it is only in recent years that publications like the Open
Science Collaboration [@opensciencecollaboration2015], Many Labs
Collaborations [@klein2018; @ebersole2020; @ebersole2016;
@klein2022; @mathur2020; @buttrick2020; @skorb2020] or the first papers
from the Psychological Science Accelerator [@moshontz2018; @jones2021;
@dorison2022; @wang2021; @legate2022; @bago2022]. Generally, the researcher incentive for replication was low for three reasons. First, journals often prioritize “novel” or new results which led to rejection of replication manuscripts and publication bias [@hubbard1997; @franco2014; @nosek2012]. Second, the
"failure" to replicate was often placed on the replication team as "bad
science" rather than a careful consideration of publication biases and
(potential) questionable research practices [@ioannidis2015;
@maxwell2015; @klein2022]. Last, why should someone want to spend time and
resources on an answer we already "know" [@isager2021; @isager2021a]?

@@

The credibility movement has been mirrored by the calls for
diversification or de-WEIRDing (e.g., Western, Educated, Industrialized,
Rich, and Democratic) scientific research [@henrich2010; @newson2021;
@rad2018] by improving representation in research samples. Like the
large-scale studies in Physics [@aphilos2021; @castelnovo2018] and
Biology [@collins2003], the Social Sciences struggle to represent the
breadth of humanity across both researcher and population
characteristics. Now, grassroots organizations, such as the
Psychological Science Accelerator [@moshontz2018], ManyBabies
(<https://manybabies.github.io/>), NutNet (<https://nutnet.org/>), and
DRAGNet (<https://dragnetglobal.weebly.com/>) can begin to tackle these
issues by recruiting research labs from all over the globe to provide
diversity in geographic, linguistic, and researcher representation.
Publications have examined the global understanding of morality, face
processing, COVID-19 information signaling, and more [@dorison2022;
@wang2021; @legate2022; @bago2022; @jones2021; @vanbavel2022]. While
these organizations and one-time groups for BTS studies have provided an
incredible wealth of data for the scientific community, we do not yet
know exactly *who* is involved with, and benefits from, the BTS and
credibility movement. Publications on BTS generally explore challenges,
lessons learned, and the need for BTS [@forscher2020; @coles2022].

@@

```{r fig-pub-time, include = TRUE, fig.cap="Number of big-team science publications separated by four large subject areas across years. All four subject areas show an exponential number of publications in the last decade. ", fig.width=10, fig.height=6}
ggplot(DF$pub_time_count %>% 
         filter(SubjectArea != "Multidisciplinary") %>% 
         filter(Year < 2023), 
       aes(Year, count, color = SubjectArea)) + 
  theme_classic(base_size = 15) + 
  geom_point() + 
  scale_color_brewer(palette = "Spectral", name = "Subject Area") + 
  ylab("Number of Articles Per Year") + 
  xlab("Publication Year") + 
  theme(legend.position = "bottom")

@@

Publication interest area was summarized by the four large subject areas
creating a word cloud plot of the total number of publications within
the ASJCs. Figure \@ref(fig:fig-clouds) displays that the Health
Sciences tends to publish within medicine and oncology, with a
corresponding focus of cancer research and genetics for the Life
Sciences. The Physical Sciences is mostly dominated by physics research,
chemistry, and ecology. The BTS publications in the Social Sciences are
mostly within psychology, education, and health.

@@

The total number of unique authors across all publications was
`r total_authors`. The mean number of authors per publication was *M* =
`r avg_authors` (*SD* = `r sd_authors`, *Med* = `r med_authors`) with a
range of `r min_authors` to `r max_authors`. The median and average
number of authors by subject area are displayed in Figure
\@ref(fig:fig-author-year). In general, the average and median number of
authors increased over time, with the exception of the skew in the
Physical Sciences. Interestingly, the effect in the Physical Sciences
appears to be declining toward the general trends seen in other areas in
the last few decades.

@@

All values for this analyses were different from zero. The slopes for
the average career length were negative for all four subject areas,
indicating a trend toward younger scientist involvement over time for
each area, with the strongest effect in the Physical Sciences. The
coefficient for variability in career length was also negative for each
of the four subject areas with the highest in the Physical Sciences and
lowest in the Life Sciences. This result indicates a decrease in the
variability of career lengths over time, likely from two sources: 1)
more publications with more authors, thus, lowering variance
estimations, and 2) more young scholars overall. The effect sizes for
this analysis were surprisingly large ranging from $R^2$ to .25 to .47.
All values and their confidence intervals can be found on our OSF page.

@@

We used the same analyses described in the career length section to
analyze trends over time. An increasing slope over time indicates that
individuals who are publishing more are more represented in BTS over
time (i.e., increasing numbers of scholars with higher publication
rates), while a negative slope indicates more researchers with less
publications. A positive slope for the standard deviation of publication
metrics indicates increasing variance over time (i.e., more diversity in
the individual publication rates), while a negative slope would indicate
less diversity in researchers over time. While publication rates do not
represent value as a researcher, they are often used in hiring and
promotion decisions, and we used this variable as a proxy to gauge the
diversity in scholars represented in big teams. As shown in Figure
\@ref(fig:fig-heatmap) publication metrics were generally negative for
the average publication metrics, indicating more scholars over time with
lower numbers of publications with the strongest effects in Health and
Social Sciences. The variability of publication counts was not
significant for the Life Sciences but was negative for the Physical
Sciences (less variability over time) and positive for Social and Health
Sciences (more variability and over time). This result indicates that
the Physical Sciences are trending toward scholars with less
publications but also less diverse in number of publications, while the
Health and Social Sciences see more diversity in publication counts and
less published scholars overall.

@@

Author geopolitical region is displayed in Figure
\@ref(fig:fig-map-both). Big team publications appear to be lead by
North America and Western Europe, while all publications are lead by
North America and East Asia. To understand the change in representation
diversity, we examined if the number of regions in a publication is
predicted by the year of publication. Increasing diversity would be
represented by a positive slope, while decreasing diversity would be
represented by a negative slope. As shown in Figure
\@ref(fig:fig-heatmap), the Physical Sciences do not show a trend of
change in representation, while all other sciences showed a positive
effect increasing in the number of geopolitical regions authors
represent on publications.

@@

In this investigation, we explored the publication rates, areas, and
researchers involved in big team science publications. Over a
half-million articles were published in nearly 15,000 journals since
1970 that qualified as big team science articles with at least ten
authors and ten different affiliations. The areas of publication aligned
to cancer and genetics research in medicine and oncology for Health and
Life Sciences, physics and chemistry for the Physical Sciences, and
psychology for the Social Sciences All areas of research show an
explosion in growth in the number of publications and the number of
author included on manuscripts, replicating previous investigations into
this topic area [@sinatra2015; @wuchty2007; @hunter2008].

@@

The variability in the types of researchers involved in publications
also decreased across time in most areas of science with a decrease in
variability for career length. As mention, an increase in early career
researchers and numbers of publications could explain this effect
mathematically, potentially with other social influences mentioned
above. The variability in number of publications is decreasing in the
Physical Sciences, mirroring the career length results, but the opposite
effect was found in the Health and Social Sciences. We see no clear
reason why career variability would decrease while the variability in
the number of publications would increase. The effect sizes for career
length were much larger than the effects for number of publications. One
speculation is the increasing requirements for a competitive faculty
role application. Given the limited number of positions, one potential
way to distinguish their application would be a larger number
publications in their early career [@caplow2017; @kyvik2003].

The number of geopolitical entities for researcher affiliation is
increasing over time, showing the results of globalization and the
ability to connect across time zones and cultures [@xie2014]. While our
definition of big team science required at least ten different
institutional affiliations, we did not filter papers by geopolitical
region, and thus, a manuscript could rely solely on institutions within
a single country. The Physical Sciences did not show an increase in
diversity of regions represented, however, it could be argued that the
development of large research centers like CERN forced earlier diversity
than other sciences (i.e., because CERN specifically recruited
scientists from sponsoring nations). The Life, Health, and Social
Sciences saw an increase in the number of regions represented with the
highest increase in the Social Science. This result likely corresponds
with an increased interest in big team science publications in
psychology [@coles2022; @forscher2020], and the desire to diversify the
populations represented in psychological research [@henrich2010;
@newson2021].
